---
title: "Hands-on Exercise 2"
subtitle: "Beyond ggplot2 Fundamentals"
author: "Akshaya Vijayakumar Sivakami"
date: "17 January 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Overview

This exercise explores several ggplot2 extensions that enhance the elegance and effectiveness of statistical graphics. The objectives are to:

-   Control the placement of annotations on graphs using functions from the ggrepel package.

-   Create professional, publication-quality figures with functions from the ggthemes and hrbrthemes packages.

-   Combine multiple ggplot2 graphs into composite figures using the patchwork package.

# Getting Started

## Installing and Loading Required Libraries

In this exercise, in addition to **tidyverse**, four additional R packages will be used:

-   **ggrepel** – provides ggplot2 geoms that prevent overlapping text labels.

-   **ggthemes** – offers additional themes, geoms, and scales to enhance ggplot2 visualizations.

-   **hrbrthemes** – supplies typography-focused themes and components for ggplot2.

-   **patchwork** – enables the creation of composite figures by combining multiple ggplot2 plots.

The code chunk below will check whether these packages are installed and then load them into the current R environment.

```{r}
pacman::p_load(ggrepel, patchwork, hrbrthemes, ggthemes, tidyverse) 
```

## Importing Data

For the purpose of this exercise, a data file called *Exam_data* will be used. It consists of year end examination grades of a cohort of primary 3 students from a local school. It is in csv file format.

The code chunk below imports *exam_data.csv* into R environment by using `read_csv()`function of [**readr**](#0) package. **readr** is one of the tidyverse package.

```{r}
exam_data <- read_csv("Exam_data.csv")
```

```{r}
summary(exam_data)
```

There are a total of seven attributes in the *exam_data* tibble data frame. Four of them are categorical data type and the other three are in continuous data type.

-   The categorical attributes are: ID, CLASS, GENDER and RACE.

-   The continuous attributes are: MATHS, ENGLISH and SCIENCE.

# Beyond ggplot2 Annotation: ggrepel

One of the challenge in plotting statistical graph is annotation, especially with large number of data points.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| eval: true
#| fig-align: "center"
ggplot(data = exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              size = 0.5) +
  geom_label(aes(label = ID),
             hjust = .5,
             vjust = -.5) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## Code

```{r}
#| echo: true
#| eval: false
ggplot(data = exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              size = 0.5) +
  geom_label(aes(label = ID),
             hjust = .5,
             vjust = -.5) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

[ggrepel](https://ggrepel.slowkow.com/) is an extension of **ggplot2** package which provides `geoms` for **ggplot2** to repel overlapping text as in the examples.

We simply replace `geom_text()` by [`geom_text_repel()`](https://ggrepel.slowkow.com/reference/geom_text_repel.html) and `geom_label()` by [`geom_label_repel`](https://ggrepel.slowkow.com/reference/geom_text_repel.html).

## Working with ggrepel

::: panel-tabset
### Plot

```{r}
#| echo: false
#| eval: true
#| fig-align: "center"
ggplot(data=exam_data, aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5) +  
  geom_label_repel(aes(label = ID), fontface = "bold") +
  coord_cartesian(xlim=c(0,100), ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

### Code

```{r}
#| echo: true
#| eval: false
ggplot(data=exam_data, aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5) +  
  geom_label_repel(aes(label = ID), fontface = "bold") +
  coord_cartesian(xlim=c(0,100), ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

# Beyond ggplot2 Themes

ggplot2 comes with eight [built-in themes](https://ggplot2.tidyverse.org/reference/ggtheme.html), they are: `theme_gray()`, `theme_bw()`, `theme_classic()`, `theme_dark()`, `theme_light()`, `theme_linedraw()`, `theme_minimal()`, and `theme_void()`.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| eval: true
#| fig-align: "center"
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=20, boundary = 100, color="grey25", fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores") 
```

## Code

```{r}
#| echo: true
#| eval: false
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=20, boundary = 100, color="grey25", fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores") 
```
:::

## Working with ggtheme package

[**ggthemes**](https://cran.r-project.org/web/packages/ggthemes/index.html) provides [‘ggplot2’ themes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) that replicate the look of plots by Edward Tufte, Stephen Few, [Fivethirtyeight](https://fivethirtyeight.com/), [The Economist](https://www.economist.com/graphic-detail), ‘Stata’, ‘Excel’, and [The Wall Street Journal](https://www.pinterest.com/wsjgraphics/wsj-graphics/), among others.

In the example below, *The Economist* theme is used.

::: panel-tabset
### Plot

```{r}
#| echo: false
#| eval: true
#| fig-align: "center"
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```

### Code

```{r}
#| echo: true
#| eval: false
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```
:::

It also provides some extra geoms and scales for ‘ggplot2’. Consult this [vignette](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) to learn more.

## Working with hrbthems package

[**hrbrthemes**](https://cran.r-project.org/web/packages/hrbrthemes/index.html) package provides a base theme that focuses on typographic elements, including where various labels are placed as well as the fonts that are used.

::: panel-tabset
### Plot

```{r}
#| echo: false
#| eval: true
#| fig-align: "center"
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```

### Code

```{r}
#| echo: true
#| eval: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```
:::

The second goal centers around productivity for a production workflow. In fact, this “production workflow” is the context for where the elements of hrbrthemes should be used. Consult [this vignette](https://hrbrmstr.github.io/hrbrthemes/articles/why_hrbrthemes.html) to learn more.

::: panel-tabset
### Plot

```{r}
#| echo: false
#| eval: true
#| fig-align: "center"
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

### Code

```{r}
#| echo: true
#| eval: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```
:::

What can we learn from the code chunk above?

-   axis_title_size argument is used to increase the font size of the axis title to 18,

-   base_size argument is used to increase the default axis label to 15,

-   and grid argument is used to remove the x-axis grid lines.

# Beyond Single Graph

It’s often necessary to use multiple graphs to convey a compelling visual story. Several **ggplot2** extensions offer functions for combining multiple plots into a single composite figure. In this section, you’ll learn how to create a composite plot by integrating several graphs. Let’s begin by generating three statistical graphics using the code chunk below.

First, a histogram showing the distribution of Maths scores is plotted.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| eval: true
#| fig-align: "center"
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")
p1
```

## Code

```{r}
#| echo: true
#| eval: false
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")
```
:::

Next, a histogram showing the distribution of English scores is plotted.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| eval: true
#| fig-align: "center"
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")
p2
```

## Code

```{r}
#| echo: true
#| eval: false
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")
```
:::

Lastly, a scatterplot for English score versus Maths score by as shown below.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| eval: true
#| fig-align: "center"
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
p3
```

## Code

```{r}
#| echo: true
#| eval: false
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

## Creating Composite Graphics: pathwork methods

There are several ggplot2 extension’s functions support the needs to prepare composite figure by combining several graphs such as [`grid.arrange()`](https://cran.r-project.org/web/packages/gridExtra/vignettes/arrangeGrob.html) of **gridExtra** package and [`plot_grid()`](https://wilkelab.org/cowplot/reference/plot_grid.html) of [**cowplot**](https://wilkelab.org/cowplot/index.html) package. In this section, I am going to shared with you an ggplot2 extension called [**patchwork**](https://patchwork.data-imaginist.com/) which is specially designed for combining separate ggplot2 graphs into a single figure.

Patchwork package has a very simple syntax where we can create layouts super easily. Here’s the general syntax that combines:

-   Two-Column Layout using the Plus Sign +.

-   Parenthesis () to create a subplot group.

-   Two-Row Layout using the Division Sign `/`

## Combining two ggplot2 graphs

Figure in the tabset below shows a composite of two histograms created using patchwork. Note how simple the syntax used to create the plot!

::: panel-tabset
### Plot

```{r}
#| echo: false
#| eval: true
#| fig-align: "center"
p1 + p2
```

### Code

```{r}
#| echo: true
#| eval: false
p1 + p2
```
:::

## Combining three ggplot2 graphs

More complex composite can be plotted by using appropriate operators. For example, the composite figure below is plotted by using:

-   “/” operator to stack two ggplot2 graphs,

-   “\|” operator to place the plots beside each other,

-   “()” operator the define the sequence of the plotting.

::: panel-tabset
### Plot

```{r}
#| echo: false
#| eval: true
#| fig-width: 10
#| fig-height: 7
#| fig-align: "center"
(p1 / p2) | p3
```

### Code

```{r}
#| echo: true
#| eval: false
(p1 / p2) | p3
```
:::

To learn more about, refer to [Plot Assembly](https://patchwork.data-imaginist.com/articles/guides/assembly.html).

## Creating a composite figure with tag

In order to identify subplots in text, patchwork also provides auto-tagging capabilities as shown in the figure below.

::: panel-tabset
### Plot

```{r}
#| echo: false
#| eval: true
#| fig-width: 10
#| fig-height: 7
#| fig-align: "center"
((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')
```

### Code

```{r}
#| echo: true
#| eval: false
((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')
```
:::

## Creating figure with insert

Beside providing functions to place plots next to each other based on the provided layout. With [`inset_element()`](https://patchwork.data-imaginist.com/reference/inset_element.html) of **patchwork**, we can place one or several plots or graphic elements freely on top or below another plot.

::: panel-tabset
### Plot

```{r}
#| echo: false
#| eval: true
#| fig-width: 10
#| fig-height: 7
#| fig-align: "center"
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```

### Code

```{r}
#| echo: true
#| eval: false
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```
:::

## Creating a composite figure by using patchwork and ggtheme

Figure below is created by combining patchwork and theme_economist() of ggthemes package discussed earlier.

::: panel-tabset
### Plot

```{r}
#| echo: false
#| eval: true
#| fig-width: 15
#| fig-height: 7
#| fig-align: "center"
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```

### Code

```{r}
#| echo: true
#| eval: false
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```
:::

# References

-   Kam, T.S. (2023). [Beyond ggplot2 Fundamentals](https://r4va.netlify.app/chap02).

-   [Patchwork R package goes nerd viral](https://www.littlemissdata.com/blog/patchwork)

-   [ggrepel](https://ggrepel.slowkow.com/)

-   [ggthemes](https://ggplot2.tidyverse.org/reference/ggtheme.html)

-   [hrbrthemes](https://cinc.rud.is/web/packages/hrbrthemes/)

-   [ggplot tips: Arranging plots](https://albert-rapp.de/post/2021-10-28-extend-plot-variety/)

-   [ggplot2 Theme Elements Demonstration](https://henrywang.nl/ggplot2-theme-elements-demonstration/)

-   [ggplot2 Theme Elements Reference Sheet](https://isabella-b.com/blog/ggplot2-theme-elements-reference/)
