---
title: "Hands-on Exercise 5-A"
subtitle: "Creating Ternary Plot with R"
author: "Akshaya Vijayakumar Sivakami"
date: "9 February 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Overview

The distribution and variability of three-part compositional data can be shown using ternary charts. (For instance, the percentage of young, economically active, and elderly people, or the amount of clay, silt, and sand in the soil.) A triangle with sides scaled from 0 to 1 is what it displays. One of the three elements is represented by each side. Plotting a point involves making sure that a line drawn perpendicular to each leg of the triangle intersects at the point's component values.

This hands-on focuses on how to use R to programmatically create ternary plots for visualising and understanding Singapore's demographic structure. It consists of four steps:

-   Install and launch tidyverse and ggtern packages.

-   Derive three new measures using mutate() function of dplyr package.

-   Build a static ternary plot using ggtern() function of ggtern package.

-   Build an interactive ternary plot using plot-ly() function of Plotly R package.

# Installing and launching R packages

For this exercise, two main R packages will be used in this hands-on exercise, they are:

-   [**ggtern**](http://www.ggtern.com/), a ggplot extension specially designed to plot ternary diagrams. The package will be used to plot static ternary plots.

-   [**Plotly R**](https://plot.ly/r/), an R package for creating interactive web-based graphs via plotly’s JavaScript graphing library, plotly.js . The **plotly R** libary contains the *ggplotly* function, which will convert **ggplot2** figures into a Plotly object.

It is also important to ensure that selected **tidyverse** family packages namely: **readr**, **dplyr** and **tidyr** are also installed and loaded.

The code chunks below will accomplish the task.

```{r}
pacman::p_load(plotly, ggtern, tidyverse)
```

# Data Preparation

## The data

For the purpose of this hands-on exercise, the [Singapore Residents by Planning AreaSubzone, Age Group, Sex and Type of Dwelling, June 2000-2018](https://www.singstat.gov.sg/find-data/search-by-theme/population/geographic-distribution/latest-data) data will be used. The dataset is called *respopagsex2000to2018_tidy.csv* and is in csv file format.

## Importing Data

To important *respopagsex2000to2018_tidy.csv* into R, ***read_csv()*** function of **readr** package will be used.

```{r}
pop_data <- read_csv("data/respopagsex2000to2018_tidy.csv") 
```

## Preparing the Data

Next, use the mutate() function of dplyr package to derive three new measures, namely: young, active, and old.

```{r}
#Deriving the young, economy active and old measures
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)
```

# Plotting Ternary Diagram with R

## Plotting a static ternary diagram

Use ggtern() function of ggtern package to create a simple ternary plot.

```{r}
#Building the static ternary plot
#| fig-align: "center"
ggtern(data=agpop_mutated,aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point()
```

```{r}
#| fig-align: "center"
#Building the static ternary plot
ggtern(data=agpop_mutated, aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point() +
  labs(title="Population structure, 2015") +
  theme_rgbw()
```

## Plotting an interative ternary diagram
The code below create an interactive ternary plot using plot_ly() function of Plotly R.

```{r}
#| fig-align: "center"
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# reusable function for axis formatting
axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

# Initiating a plotly visualization 
plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  )
```

# References

-   Kam. T.S. (2023). [Creating Ternary Plot with R](https://r4va.netlify.app/chap13)
