---
title: "Hands-on Exercise 5-D"
subtitle: "Visualising Multivariate Analysis with Parallel Coordinates Plot"
author: "Akshaya Vijayakumar Sivakami"
date: "10 February 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Overview

A parallel coordinates plot is a specialized visualization technique engineered for exploring and analyzing high-dimensional, multivariate numerical data. It excels at revealing complex relationships and correlations across multiple variables simultaneously—such as the various socioeconomic factors that aggregate into a "Happiness Index."

Developed by Alfred Inselberg in the 1970s, this method is a staple in scientific and academic research but remains less common in general business reporting. Data visualization expert Stephen Few (2006) highlights its specific utility:

*"This certainly isn’t a chart that you would present to the board of directors... the strength of parallel coordinates isn’t in their ability to communicate some truth in the data to others, but rather in their ability to bring meaningful multivariate patterns and comparisons to light when used interactively for analysis."*

In a business context, it is particularly effective for characterizing clusters identified during customer segmentation, allowing analysts to see exactly how different segments behave across various metrics.

This hands-on exercise aims to give exposure to:

-   plotting statistic parallel coordinates plots by using [`ggparcoord()`](https://ggobi.github.io/ggally/reference/ggparcoord.html) of [**GGally**](https://r4va.netlify.app/chap15) package, and

-   plotting interactive parallel coordinates plots by using [**parallelPlot**](https://cran.r-project.org/web/packages/parallelPlot/) package.

# Installing and Launching R Packages

For this exercise, the **GGally**, **parallelPlot** and **tidyverse** packages will be used.

The code chunk below is used to install and load the packages in R.

```{r}
pacman::p_load(GGally, parallelPlot, tidyverse)
```

# Data Preparation

In this hands-on exercise, the [World Happinees 2018](https://www.worldhappiness.report/ed/2018/) data will be used. The original data set is in Microsoft Excel format. It has been extracted and saved in csv file called **WHData-2018.csv**.

In the code chunk below, `read_csv()` of **readr** package is used to import *WHData-2018.csv* into R and save it into a tibble data frame object called *wh*.

```{r}
wh <- read_csv("data/WHData-2018.csv")
```

# Plotting Static Parallel Coordinates Plot

This section explores how to plot static parallel coordinates plot by using [`ggparcoord()`](https://ggobi.github.io/ggally/reference/ggparcoord.html) of **GGally** package.

## Plotting a simple parallel coordinates

Code chunk below shows a typical syntax used to plot a basic static parallel coordinates plot by using `ggparcoord()`.

```{r}
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 10
ggparcoord(data = wh, 
           columns = c(7:12))
```

Notice that only two argument namely `data` and `columns` are used. `Data` argument is used to map the data object (i.e. *wh*) and `columns` is used to select the columns for preparing the parallel coordinates plot.

## Plotting a parallel coordinates with boxplot

The basic parallel coordinates failed to reveal any meaning understanding of the World Happiness measures. This section explores how to makeover the plot by using a collection of arguments provided by `ggparcoord()`.

```{r}
#| fig-align: "center"
#| fig-width: 13
#| fig-height: 10
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plot of World Happines Variables")
```

**Things to learn from the code chunk above:**

-   `groupColumn` argument is used to group the observations (i.e. parallel lines) by using a single variable (i.e. Region) and colour the parallel coordinates lines by region name.

-   `scale` argument is used to scale the variables in the parallel coordinate plot by using `uniminmax` method. The method univariately scale each variable so the minimum of the variable is zero and the maximum is one.

-   `alphaLines` argument is used to reduce the intensity of the line colour to 0.2. The permissible value range is between 0 to 1.

-   `boxplot` argument is used to turn on the boxplot by using logical `TRUE`. The default is `FALSE`.

-   `title` argument is used to provide the parallel coordinates plot a title.

## Parallel coordinates with facet

Since `ggparcoord()` is developed by extending **ggplot2** package, some of the ggplot2 functions can be combined when plotting a parallel coordinates plot.

In the code chunk below, `facet_wrap()` of **ggplot2** is used to plot 10 small multiple parallel coordinates plots. Each plot represent one geographical region such as East Asia.

```{r}
#| fig-align: "center"
#| fig-width: 14
#| fig-height: 9
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)
```

One of the aesthetic defect of the current design is that some of the variable names overlap on x-axis.

## Rotating x-axis text label

To make the x-axis text label easy to read, the labels are rotated by 30 degrees. The axis text labels can be rotated using `theme()` function in ggplot2 as shown in the code chunk below.

```{r}
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 8
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30))
```

::: callout-note
To rotate x-axis text labels, we use `axis.text.x` as argument to `theme()` function. The `element_text(angle = 30)` is specified to rotate the x-axis text by an angle 30 degree.
:::

## Adjusting the rotated x-axis text label

Rotating x-axis text labels to 30 degrees makes the label overlap with the plot and this can be avoided by adjusting the text location using `hjust` argument to theme’s text element with `element_text()`. `axis.text.x` can be used to change the look of x-axis text.

```{r}
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 8
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30, hjust=1))
```

# Plotting Interactive Parallel Coordinates Plot: parallelPlot methods

[**parallelPlot**](https://cran.r-project.org/web/packages/parallelPlot/) is an R package specially designed to plot a parallel coordinates plot by using ‘htmlwidgets’ package and [d3.js](https://d3js.org/). This section dives into using functions provided in parallelPlot package to build interactive parallel coordinates plot.

## The basic plot

The code chunk below plot an interactive parallel coordinates plot by using `parallelPlot()`.

```{r}
wh <- wh %>%
  select("Happiness score", c(7:12))
parallelPlot(wh,
             width = 320,
             height = 250)
```

::: callout-note
Some of the axis labels are too long. The next steps details how to overcome this problem.
:::

## Rotate axis label

In the code chunk below, `rotateTitle` argument is used to avoid overlapping axis labels.

```{r}
parallelPlot(wh,
             rotateTitle = TRUE)
```

A standout interactive feature of **parallelPlot** is the ability to apply a color gradient on demand. By simply clicking a variable, like the **Happiness score,** the default solid blue lines transform into a multi-toned blue color scheme. This intensity shift makes it much easier to track how different values for that specific variable behave across the rest of the plot.

## Changing the colour scheme

The default blue colour scheme can be changed by using `continousCS` argument as shown in the code chunk below.

```{r}
parallelPlot(wh,
             continuousCS = "YlOrRd",
             rotateTitle = TRUE)
```

## Parallel coordinates plot with histogram

In the code chunk below, `histoVisibility` argument is used to plot histogram along the axis of each variables.

```{r}
histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)
```

# References

-   Kam, T.S. (2023). [Visual Multivariate Analysis with Parallel Coordinates Plot](https://r4va.netlify.app/chap15)

-   [*ggparcoord()*](http://ggobi.github.io/ggally/reference/ggparcoord.html) of [**GGally**](http://ggobi.github.io/ggally/index.html) package

-   [**parallelPlot**](https://cran.r-project.org/web/packages/parallelPlot/parallelPlot.pdf)
