---
title: "Hands-on Exercise 7-C"
subtitle: "Analytical Mapping"
author: "Akshaya Vijayakumar Sivakami"
date: "3 March 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Overview

This exercise focuses on using appropriate R methods to plot analytical maps. By the end of this exercise, the appropriate functions of **tmap** and **tidyverse** will be used to perform the following tasks:

-   Importing geo-spatial data in rds format into the R environment

-   Creating cartographic-quality choropleth maps through the use of appropriate ***tmapfunctions***

-   Creating a rate map

-   Creating a percentile map

-   Creating a boxmap

# Getting Started

## Installing and launching R packages

Before starting, install and load **sf**, **tmap** and **tidyverse** packages into R environment.

```{r}
pacman::p_load(tmap, tidyverse, sf)
```

## Importing data

For the purpose of this hands-on exercise, a prepared data set called *NGA_wp.rds* will be used. The data set is a polygon feature data.frame providing information on water point of Nigeria at the LGA level.

```{r}
NGA_wp <- read_rds("data/NGA_wp.rds")
```

# Basic Choropleth Mapping

Here, a choropleth map showing the distribution of non-function water point by LGA is plotted.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-align: 'center'
p1 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "wp_functional",
             fill.scale = tm_scale_intervals(
               style = "equal",
               n = 10,
               values = "brewer.blues"),
             fill.legend = tm_legend(
               position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Distribution of functional water point by LGAs") 

p2 <- tm_shape(NGA_wp) + 
  tm_polygons(fill = "total_wp", 
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1, 
             fill_alpha = 1) + 
  tm_title("Distribution of total  water point by LGAs")

tmap_arrange(p2, p1, nrow = 1)
```

# Choropleth Map for Rates

In much of the readings, the importance of mapping rates rather than counts has been emphasized, and this is due to the fact that water points are not equally distributed across space. Consequently, if the number of water points in a given area is not accounted for, what is mapped is the total size of water points rather than the intended topic of interest.

## Deriving Proportion of Functional Water Points and Non-Functional Water Points

Here, the proportion of functional water points and the proportion of non-functional water points in each LGA will be mapped. In the following code chunk, `mutate(`) from **dplyr** package is used to derive two fields, namely *pct_functional* and *pct_nonfunctional*.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
NGA_wp <- NGA_wp %>% mutate(pct_functional = wp_functional/total_wp) %>% mutate(pct_nonfunctional = wp_nonfunctional/total_wp)
```

## Plotting map of rate

The code chunk below plots a choropleth map showing the distribution of percentage functional water point by LGA.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-align: 'center'
#| fig-width: 10
tm_shape(NGA_wp) +
  tm_polygons("pct_functional",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) + 
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Rate map of functional water point by LGAs")
```

# Extreme Value Maps

Extreme value maps represent a variation of conventional choropleth maps, in which the classification scheme is designed to emphasize extreme values at both the lower and upper ends of the scale, aiming to identify outliers. These maps were developed in the context of spatializing exploratory data analysis (EDA), that is, incorporating spatial components into the methods commonly applied in non-spatial EDA (Anselin, 1994).

## Percentile Map

A percentile map is a specific type of quantile map that uses six defined categories: 0–1%, 1–10%, 10–50%, 50–90%, 90–99%, and 99–100%. The corresponding breakpoints can be obtained in base R using the quantile function, by specifying a vector of cumulative probabilities as c(0, .01, .1, .5, .9, .99, 1). It should be noted that both the starting and ending points must be included.

### Data Preparation

First, the records with NA are excluded. Then, customised classification is created and values are extracted by using the code chunk below.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
NGA_wp <- NGA_wp %>%
  drop_na()

percent <- c(0,.01,.1,.5,.9,.99,1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)
quantile(var[,1], percent)
```

When variables are extracted from an sf data.frame, the geometry is extracted as well. For mapping and spatial manipulation, this is the expected behavior, but many base R functions cannot deal with the geometry. Specifically, the `quantile()` gives an error. As a result `st_set_geomtry(NULL)` is used to drop geomtry field.

### Why writing functions?

Writing a function has three big advantages over using copy-and-paste:

-   A function can be given an evocative name that makes the code easier to understand.

-   As requirements change, the code needs to be updated only in one place, instead of many.

-   It eliminates the chance of making incidental mistakes when copying and pasting (i.e. updating a variable name in one place, but not in another).

Source: [Chapter 19: Functions](https://r4ds.had.co.nz/functions.html#functions) of **R for Data Science**.

### Creating the get.var function

Firstly, an R function is written as shown below to extract a variable (i.e. *wp_nonfunctional*) as a vector out of an sf data.frame.

-   arguments:

    -   vname: variable name (as character, in quotes)

    -   df: name of sf data frame

-   returns:

    -   v: vector with values (without a column name)

```{r}
#| code-fold: true
#| code-summary: "Show the code"
get.var <- function(vname,df) {
  v <- df[vname] %>% 
    st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

### A percentile mapping function

Next, a percentile mapping function is written by using the code chunk below.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
percentmap <- function(vnam, df, legtitle=NA, mtitle="Percentile Map"){
  percent <- c(0,.01,.1,.5,.9,.99,1)
  var <- get.var(vnam, df)
  bperc <- quantile(var, percent)
  tm_shape(df) +
  tm_polygons() +
  tm_shape(df) +
     tm_polygons(vnam,
             title=legtitle,
             breaks=bperc,
             palette="Blues",
          labels=c("< 1%", "1% - 10%", "10% - 50%", "50% - 90%", "90% - 99%", "> 99%"))  +
  tm_borders() +
  tm_layout(main.title = mtitle, 
            title.position = c("right","bottom"))
}
```

### Test drive the percentile mapping function

Finally, the percentile map is plottes using the percentile mapping function.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-align: 'center'
#| fig-width: 10
percentmap("total_wp", NGA_wp)
```

Note that this is just a bare bones implementation. Additional arguments such as the title, legend positioning just to name a few of them, could be passed to customise various features of the map.

## Box map

Essentially, a box map is an enhanced version of a quartile map, incorporating additional lower and upper categories. When lower outliers are present, the starting point of the breaks is set to the minimum value, while the second break corresponds to the lower fence. Conversely, if no lower outliers exist, the breaks begin at the lower fence, with the second break representing the minimum value—resulting in no observations falling within the interval between the lower fence and the minimum value.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-align: 'center'
#| fig-width: 10
ggplot(data = NGA_wp,
       aes(x = "",
           y = wp_nonfunctional)) +
  geom_boxplot()
```

-   Displaying summary statistics on a choropleth map by using the basic principles of boxplot.

-   To create a box map, a custom breaks specification will be used. However, there is a complication. The break points for the box map vary depending on whether lower or upper outliers are present.

### Creating the boxbreaks function

The code chunk below is an R function that creating break points for a box map.

-   arguments:

    -   v: vector with observations

    -   mult: multiplier for IQR (default 1.5)

-   returns:

    -   bb: vector with 7 break points compute quartile and fences

```{r}
#| code-fold: true
#| code-summary: "Show the code"
boxbreaks <- function(v,mult=1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  # initialize break points vector
  bb <- vector(mode="numeric",length=7)
  # logic for lower and upper fences
  if (lofence < qv[1]) {  # no lower outliers
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) { # no upper outliers
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}
```

### Creating the get.var function

The code chunk below is an R function to extract a variable as a vector out of an sf data frame.

-   arguments:

    -   vname: variable name (as character, in quotes)

    -   df: name of sf data frame

-   returns:

    -   v: vector with values (without a column name)

```{r}
#| code-fold: true
#| code-summary: "Show the code"
get.var <- function(vname,df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

### Test drive the newly created function

Let’s test the newly created function.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-align: 'center'
var <- get.var("wp_nonfunctional", NGA_wp) 
boxbreaks(var)
```

### Boxmap function

The code chunk below is an R function to create a box map. - arguments: - vnam: variable name (as character, in quotes) - df: simple features polygon layer - legtitle: legend title - mtitle: map title - mult: multiplier for IQR - returns: - a tmap-element (plots a map)

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-align: 'center'
#| fig-width: 10
boxmap <- function(vnam, df, 
                   legtitle=NA,
                   mtitle="Box Map",
                   mult=1.5){
  var <- get.var(vnam,df)
  bb <- boxbreaks(var)
  tm_shape(df) +
    tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,title=legtitle,
             breaks=bb,
             palette="Blues",
          labels = c("lower outlier", 
                     "< 25%", 
                     "25% - 50%", 
                     "50% - 75%",
                     "> 75%", 
                     "upper outlier"))  +
  tm_borders() +
  tm_layout(main.title = mtitle, 
            title.position = c("left",
                               "bottom"))
}

tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)

```

# References

-   Kam, T.S. (2023). [Analytical Mapping](https://r4va.netlify.app/chap23)
