---
title: "Hands-on Exercise 7-B"
subtitle: "Visualising Geospatial Point Data"
author: "Akshaya Vijayakumar Sivakami"
date: "3 March 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Overview

Proportional symbol maps (also referred to as graduated symbol maps) are classified as maps in which differences in the magnitude of a discrete, abruptly changing phenomenon, such as counts of people, are represented through the visual variable of size.

As with choropleth maps, classed or unclassed versions of these maps may be created. The classed versions are referred to as range-graded or graduated symbols, while the unclassed versions are called proportional symbols, in which the areas of the symbols are made proportional to the values of the mapped attribute.

In this hands-on exercise, the creation of a proportional symbol map displaying the number of wins by Singapore Pools’ outlets will be carried out using the R package **tmap**.

# Installing and Launching R Packages

Before starting, it is important to ensure that **tmap** package of R and other related R packages have been installed and loaded into R.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

# Geo-spatial Data Wrangling

The data set use for this hands-on exercise is called *SGPools_svy21*. The data is in csv file format.

Figure below shows the first 15 records of the data set. It consists of seven columns. The XCOORD and YCOORD columns are the x-coordinates and y-coordinates of SingPools outlets and branches. They are in [Singapore SVY21 Projected Coordinates System](https://www.sla.gov.sg/sirent/CoordinateSystems.aspx).

## Data Import and Preparation

The code chunk below uses `read_csv()` function of **readr** package to import *SGPools_svy21.csv* into R as a tibble data frame called *sgpools*.

```{r}
sgpools <- read_csv("data/SGPools_svy21.csv")
```

After importing the data file into R, it is important to examine if the data file has been imported correctly.The code chunk below uses `list()` to do this.

```{r}
list(sgpools) 
```

## Creating a sf data frame from an aspatial data frame

The code chunk below converts *sgpools* data frame into a simple feature data frame by using `st_as_sf()` of **sf** packages.

```{r}
sgpools_sf <- st_as_sf(sgpools, 
                       coords = c("XCOORD", "YCOORD"),
                       crs= 3414)
```

Things to learn from the arguments above:

-   The *`coords`* argument requires providing the column name of the x-coordinates first then followed by the column name of the y-coordinates.

-   The *`crs`* argument requires providing the coordinates system in epsg format. [EPSG: 3414](https://epsg.io/3414) is Singapore SVY21 Projected Coordinate System. Other country’s epsg code can be searched by referring to [epsg.io](https://epsg.io/).

Figure below shows the data table of *sgpools_sf*. Notice that a new column called geometry has been added into the data frame.

The basic information of the newly created *sgpools_sf* can be displayed by using the code chunk below.

```{r}
list(sgpools_sf)
```

The output shows that *sgppols_sf* is in point feature class. Its epsg ID is 3414. The bbox provides information of the extend of the geo-spatial data.

# Drawing a Symbol Map

To create an interactive proportional symbol map in R, the view mode of tmap will be used.

The code chunk below will turn on the interactive mode of tmap.

```{r}
tmap_mode("view")
```

## Interactive Point Symbol Map

Here, an interactive point symbol map is created to visualise the SGPools data.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-align: 'center'
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "red",
           size = 1,
           col = "black",
           lwd = 1)
```

## Proportional Symbol Map

To draw a proportional symbol map, a numerical variable should be assigned to the size visual attribute. Here, the variable "*Gp1Gp2Winnings"* is assigned to size visual attribute.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-align: 'center'
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "red",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```

## Changing the Colour

The proportional symbol map can be further improved by using the colour visual attribute. In the code chunk below, "*OUTLET_TYPE"* variable is used as the colour attribute variable.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-align: 'center'
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "OUTLET TYPE", 
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```

## Faceted Plots using tm_facets()

An impressive and little-know feature of **tmap**’s view mode is that it also works with faceted plots. The argument `sync` in `tm_facets()` can be used in this case to produce multiple maps with synchronised zoom and pan settings.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-align: 'center'
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "OUTLET TYPE", 
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1) + 
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```

# References

-   Kam, T.S. (2023). [Visualising Geospatial Point Data](https://r4va.netlify.app/chap22)

## All about tmap package

-   [tmap: Thematic Maps in R](https://www.jstatsoft.org/article/view/v084i06)

-   [tmap](https://cran.r-project.org/web/packages/tmap/index.html)

-   [tmap: get started!](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)

-   [tmap: changes in version 2.0](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-changes-v2.html)

-   [tmap: creating thematic maps in a flexible way (useR!2015)](http://von-tijn.nl/tijn/research/presentations/tmap_user2015.pdf)

-   [Exploring and presenting maps with tmap (useR!2017)](http://von-tijn.nl/tijn/research/presentations/tmap_user2017.pdf)

## Geo-spatial data wrangling

-   [sf: Simple Features for R](https://cran.r-project.org/web/packages/sf/index.html)

-   [Simple Features for R: StandardizedSupport for Spatial Vector Data](https://journal.r-project.org/archive/2018/RJ-2018-009/RJ-2018-009.pdf)

-   [Reading, Writing and Converting Simple Features](https://cran.r-project.org/web/packages/sf/vignettes/sf2.html)

## Data wrangling

-   [dplyr](https://dplyr.tidyverse.org/)

-   [Tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)

-   [tidyr: Easily Tidy Data with ‘spread()’ and ‘gather()’ Functions](https://cran.r-project.org/web/packages/tidyr/tidyr.pdf)
