---
title: "Hands-on Exercise 9"
subtitle: "Information Dashboard Design: R methods"
author: "Akshaya Vijayakumar Sivakami"
date: "23 March 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Overview

This hands-on exercise explores how to:

-   create bullet chart by using ggplot2

-   create sparklines by using ggplot2

-   build industry standard dashboard by using R Shiny.

# Getting Started

## Installing and Loading Required Packages

The following R packages will be used in this exercise:

-   [tidyverse](https://www.tidyverse.org/) provides a collection of functions for performing data science task such as importing, tidying, wrangling data and visualising data. It is not a single package but a collection of modern R packages including but not limited to [readr](https://readr.tidyverse.org/), [tidyr](https://tidyr.tidyverse.org/), [dplyr](https://dplyr.tidyverse.org/), [ggplot](https://ggplot2.tidyverse.org/), [tibble](https://tibble.tidyverse.org/), [stringr](https://stringr.tidyverse.org/), [forcats](https://forcats.tidyverse.org/) and [purrr](https://purrr.tidyverse.org/).

-   [lubridate](https://r4va.netlify.app/chap31) provides functions to work with dates and times more efficiently.

-   [ggthemes](https://r4va.netlify.app/chap31) is an extension of ggplot2. It provides additional [themes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) beyond the basic themes of ggplot2.

-   [gtExtras](https://jthomasmock.github.io/gtExtras/index.html) provides some additional helper functions to assist in creating beautiful tables with [gt](https://gt.rstudio.com/), an R package specially designed for anyone to make wonderful-looking tables using the R programming language.

-   [reactable](https://glin.github.io/reactable/index.html) provides functions to create interactive data tables for R, based on the [React Table](https://github.com/tanstack/table/tree/v7) library and made with [reactR](https://github.com/react-R/reactR).

-   [reactablefmtr](https://kcuilla.github.io/reactablefmtr/index.html) provides various features to streamline and enhance the styling of interactive reactable tables with easy-to-use and highly-customizable functions and themes.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
pacman::p_load(tidyverse, lubridate, ggthemes, gtExtras, reactable, reactablefmtr, gt)
```

## Importing Dataset

For this exercise, the *CoffeeChain.rds* is used.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
coffeechain <- read_rds("data/CoffeeChain.rds")
```

## Data Preparation

Before plotting the bullet chart, the Sales and Budgeted Sales are aggregated at the Product level.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`)) %>%
  ungroup()
```

## Bullet Chart in ggplot2

The code chunk below is used to plot the bullet charts using ggplot2 functions.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
ggplot(product, aes(Product, current)) + 
  geom_col(aes(Product, max(target) * 1.01),
           fill="grey85", width=0.85) +
  geom_col(aes(Product, target * 0.75),
           fill="grey60", width=0.85) +
  geom_col(aes(Product, target * 0.5),
           fill="grey50", width=0.85) +
  geom_col(aes(Product, current), 
           width=0.35,
           fill = "black") + 
  geom_errorbar(aes(y = target,
                    x = Product, 
                    ymin = target,
                    ymax= target), 
                width = .4,
                colour = "red",
                size = 1) +
  coord_flip()
```

# Plotting sparklines using ggplot2

To plot sparklines using ggplot2, a monthly product sales summary from *coffeechain* for records on or after January 1, 2013 is generated, showing only Month, Product, and the summed Sales for each month-product combination.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)
```

Next, the minimum, maximum and end of the the month sales are computed.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
mins <- group_by(sales_report, Product) %>% 
  slice(which.min(Sales))
maxs <- group_by(sales_report, Product) %>% 
  slice(which.max(Sales))
ends <- group_by(sales_report, Product) %>% 
  filter(Month == max(Month))
```

Finally, the 25 and 75 quantiles are computed.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(quart1 = quantile(Sales, 
                              0.25),
            quart2 = quantile(Sales, 
                              0.75)) %>%
  right_join(sales_report)
```

Using the *sales_report* data, the sparklines is plotted.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-align: "center"
ggplot(sales_report, aes(x=Month, y=Sales)) + 
  facet_grid(Product ~ ., scales = "free_y") + 
  geom_ribbon(data = quarts, aes(ymin = quart1, max = quart2), 
              fill = 'grey90') +
  geom_line(size=0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins, aes(label = Sales), vjust = -1) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0) +
  expand_limits(x = max(sales_report$Month) + 
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title=element_blank(), axis.text.y = element_blank(), 
        axis.ticks = element_blank(), strip.text = element_blank())
```


# Static Information Dashboard Design: gt and gtExtras methods

This section shows how to create static information dashboard by using [**gt**](https://gt.rstudio.com/) and [**gtExtras**](https://jthomasmock.github.io/gtExtras/index.html) packages.

## Plotting a Simple Bullet Chart

This sections focuses on preparing a bullet chart report by using functions of **gt** and **gtExtras** packages.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current, 
              target = target, 
              width = 60,
              palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538()
```

# sparklines: gtExtras method

Before preparing the sales report by product by using gtExtras functions, a sales summary by product and month is generated, which is then added up to total sales for each product-month combination. The code chunk below returns a grouped table with Product, Month, and Sales.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate (Month = month(Date, 
                        label = TRUE, 
                        abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()
```

It is important to note that one of the requirement of gtExtras functions is that almost exclusively they require you to pass data.frame with list columns. In view of this, code chunk below will be used to convert the report data.frame into list columns.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")
```

## Plotting CoffeeChain Sales Report

A **gt** table is created where each product has its monthly sales stored as a list, then replaces those numbers with a sparkline for each row. It shows a tiny trend line of sales by product, and `same_limit = FALSE` lets each sparkline use its own scale instead of sharing one axis range.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop") %>%
   gt() %>%
   gt_plt_sparkline('Monthly Sales',
                    same_limit = FALSE)
```

## Adding Statistics

First, summary statistics such as Max, Min, and Average are calculated.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            ) %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2)
```

## Combining the Dataframe

Next, the summary statistics are added to the table using `left_join()` to get *sales_data* table.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")

sales <- report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            )

sales_data = left_join(sales, spark)
```

## Plotting the updated Data Table

The *sales_data* table is then plotted as sparklines.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

## Combining bullet chart and sparklines

Similarly, the bullet chart and sparklined can be combined, as shown below.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(`Product`) %>%
  summarise(`Target` = sum(`Budget Sales`),
            `Actual` = sum(`Sales`)) %>%
  ungroup() 

sales_data = sales_data %>%
  left_join(bullet)

sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales') %>%
  gt_plt_bullet(column = Actual, 
                target = Target, 
                width = 28,
                palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538()
```

# Interactive Information Dashboard Design: reactable and reactablefmtr methods

This section explores how to create interactive information dashboard by using [**reactable**](https://glin.github.io/reactable/index.html) and [**reactablefmtr**](https://kcuilla.github.io/reactablefmtr/index.html) packages.

In order to build an interactive sparklines, it is neceassry to install dataui R package (if not already done). Then the package should be loaded into the R environment.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# remotes::install_github("timelyportfolio/dataui")

library(dataui)
```

## Plotting Interactive Sparklines

Similar to **gtExtras**, to plot an interactive sparklines by using **reactablefmtr** package, the list field should be prepared.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))
```

Next, [`react_sparkline`](https://kcuilla.github.io/reactablefmtr/reference/react_sparkline.html) will be used to plot the sparklines as shown below.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

## Changing the Page Size

By default the pagesize is 10. This can be changed by using the argument `defaultPageSize`.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

## Adding Points and Labels

Here, `highlight_points` argument is used to show the minimum and maximum values points and `label` argument is used to label first and last values.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
        )
    )
  )
)
```

## Adding Reference Line

To show the mean line, `statline` argument is used.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
        )
    )
  )
)
```

## Adding Bandline

Instead adding reference line, bandline can be added by using the `bandline` argument.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
        )
    )
  )
)
```

## Changing sparkline to sparkbar

Instead of displaying the values as sparklines, they can be displayed as sparkbars as well.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)
```

# References

-   Kam, T.S. (2023). [Information Dashboard Design: R methods](https://r4va.netlify.app/chap31)
